---
title: "607 final project"
author: "Jay Lee and Benson Toi"
date: "5/3/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(infer)
library(ggplot2)
library(readr)
library(usmap)
library(stringr)
library(extraoperators)
library(psych)
library(jsonlite)
library(tidyverse)
```

### Introduction


In recent years, we have heard hate crimes more often from the media like New York Times. Hate Crimes are offenses that are motivated by a particular race, religion, ethnicity, gender, age, disability, ancestry, national origin or sexual orientation. Rising hate crimes may be due to “the new normal” that was caused by the pandemic. Pandemic also cause a lot of people to lose their jobs. Furthermore, does unemployment rate, household income and education level drive hate crimes as well? Is there a relationship between the unemployment rate, household income and education level and hate crimes?

### New York Times API 

From the NYT archive API, we found hate crime related articles increase 300%. 

## load libary and apikey in backend
```{r libary, include=FALSE}
apikey <- "61r4qgmnWWFK8x2cXRj3IDGOXAmPwQzn"
```
## look at the data and pull the useful info


```{r url, echo=TRUE}
nyturl19 <- paste("https://api.nytimes.com/svc/archive/v1/2019/1.json?api-key=", apikey, sep='')
hateData19 <- fromJSON(nyturl19) #Look at what is useful
hateData19 <- hateData19 %>% 
as.data.frame() %>%
select(-response.docs.multimedia) %>%
filter(str_detect(response.docs.lead_paragraph, "hate crime"))
nyturl20 <- paste("https://api.nytimes.com/svc/archive/v1/2020/1.json?api-key=", apikey, sep='')
hateData20 <- fromJSON(nyturl20) #Look at what is useful
hateData20 <- hateData20 %>% 
as.data.frame() %>%
select(-response.docs.multimedia) %>%
filter(str_detect(response.docs.lead_paragraph, "hate crime"))
hateData19 <- hateData19 %>% 
select(response.docs.pub_date,response.docs.lead_paragraph)
hateData20 <- hateData20 %>% 
select(response.docs.pub_date,response.docs.lead_paragraph)
print(hateData19)
print(hateData20)
```
### Data

The data are from FBI, Education Attainment in the U.S.  and Southern Poverty Law Center. The FBI Uniform Crime Reporting Program collects hate crime data from law enforcement agencies. The UCR Program collects data on only prosecutable hate crimes, which make up a fraction of hate incidents (which includes non-prosecutable offenses, such as circulation of white nationalist recruitment materials on college campuses). The Southern Poverty Law Center uses media accounts and people’s self-reports to assess the situation. The Education Attainment is from U.S. Census Bureau. 

The Southern Poverty Law Center uses media accounts and people’s self-reports to assess the situation.

We want to define what is high unemployed rate and education level, so we use median as a dividing line.

If the share_unemployed_seasonal is higher than median, then the reply under high_unemployed is true, else is false.

If the PercentBachelorsOrHigher is higher than median, then the reply under high_education is true, else is false.

Here is the summary of the data:

```{r include=FALSE}
# load data
hate_urlforplot<- "https://raw.githubusercontent.com/fivethirtyeight/data/master/hate-crimes/hate_crimes.csv"
hate_urlforplot <-read.csv(hate_urlforplot)
hate_url<- "https://raw.githubusercontent.com/fivethirtyeight/data/master/hate-crimes/hate_crimes.csv"
hate_url <-read.csv(hate_url)
hate_url_benson<- "https://raw.githubusercontent.com/fivethirtyeight/data/master/hate-crimes/hate_crimes.csv"
hate_url_ds <-read.csv(hate_url_benson)
Education_rate<- "https://raw.githubusercontent.com/Benson90/607-final-project/main/csvData-2.csv"
Education_rate_ds <-read.csv(Education_rate)


# step1 pull the useful column
hate_url <- hate_url %>%
select("state","median_household_income","share_unemployed_seasonal","hate_crimes_per_100k_splc","avg_hatecrimes_per_100k_fbi")

# step2 exclude the NA
hate_url <- hate_url[complete.cases(hate_url),]

# step3 added a new column to combine both case from FBI, Southern Poverty Law Center number and education level
hate_url$hate_crimes_combine <- hate_url$hate_crimes_per_100k_splc+hate_url$avg_hatecrimes_per_100k_fbi 

hate_education_ds <- hate_url_ds %>%
  select("state","hate_crimes_per_100k_splc","avg_hatecrimes_per_100k_fbi")


hate_education_ds$avg_hatecrimes_per_100k_fbi <- hate_education_ds$avg_hatecrimes_per_100k_fbi / 100000
hate_education_ds$hate_crimes_per_100k_splc <- hate_education_ds$hate_crimes_per_100k_splc / 100000

Education_rate_ds$PercentHighSchoolOrHigher <- Education_rate_ds$PercentHighSchoolOrHigher / 100
Education_rate_ds$PercentBachelorsOrHigher <- Education_rate_ds$PercentBachelorsOrHigher / 100

joined_tibble <- left_join(hate_education_ds, Education_rate_ds, 
              by = c("state" = "State"))


# step4 get the Median which is 0.05200
summary(hate_url) 

summary(hate_education_ds)
summary(Education_rate_ds)

summary(joined_tibble)

# step5 define high and low unemployed rate by Median
hate_url$high_unemployed <-hate_url$share_unemployed_seasonal %g% 0.05200 
hate_url$high_housegoldincome<-hate_url$median_household_income %g% 54310 

# step6 share_unemployed_seasonal*100k to get unemployed count per 100k people
hate_url$share_unemployed_seasonal_100k <- hate_url$share_unemployed_seasonal*100000



```
```{r echo=FALSE}
summary(hate_url) 
```

### Exploratory data analysis

First of all, lets check the the unemployement level, hate crime level, and high education status in each states. Then go thru the data distribution , summary of unemployed rate, household income, and education level.

```{r}

#mapping 

plot_usmap(data = joined_tibble, values = "hate_crimes_per_100k_splc", color = "red") + 
  scale_fill_continuous(
    low = "white", high = "red", name = "Hate Crimes SPLC source", label = scales::comma
  ) + theme(legend.position = "right")


plot_usmap(data = joined_tibble, values = "avg_hatecrimes_per_100k_fbi", color = "red") + 
  scale_fill_continuous(
    low = "white", high = "red", name = "Hate Crimes fbi source", label = scales::comma
  ) + theme(legend.position = "right")

plot_usmap(data = hate_urlforplot, values = "median_household_income", color = "purple") + 
  scale_fill_continuous(
    low = "white", high = "purple", name = "Household income arcoss US", label = scales::comma
  ) + theme(legend.position = "right")
plot_usmap(data = hate_urlforplot, values = "share_unemployed_seasonal", color = "brown") + 
  scale_fill_continuous(
    low = "white", high = "brown", name = "unemployment arcoss US", label = scales::comma
  ) + theme(legend.position = "right")

plot_usmap(data = joined_tibble, values = "PercentBachelorsOrHigher", color = "blue") + 
  scale_fill_continuous(
    low = "white", high = "blue", name = "Bachelors Or Higher", label = scales::comma
  ) + theme(legend.position = "right")

plot_usmap(data = joined_tibble, values = "PercentHighSchoolOrHigher", color = "blue") + 
  scale_fill_continuous(
    low = "white", high = "blue", name = "HighSchool Or Higher", label = scales::comma
  ) + theme(legend.position = "right")
```

```{r echo=FALSE}
ggplot(hate_url, aes(share_unemployed_seasonal_100k)) +        
  geom_histogram(aes(y = ..density..),bins=20) +
  stat_function(fun = dnorm,
                args = list(mean = mean(hate_url$share_unemployed_seasonal_100k),
                            sd = sd(hate_url$share_unemployed_seasonal_100k)),
                col = "#1b98e0",
                size = 1)
summary(hate_url$share_unemployed_seasonal_100k)
```
Now lets check the data distribution and summary of hate crimes case. Hate crimes data is different from previous result, the data appear right skewed for me, and clearly we can see one outliers from the Now lets check the data distribution and summary of crimes case. It is different from previous result, the data appear right skewed for me, and clearly we can see one outliers from the plot.


```{r echo=FALSE}
ggplot(hate_url, aes(hate_crimes_combine)) +        
  geom_histogram(aes(y = ..density..),bins=20) +
  stat_function(fun = dnorm,
                args = list(mean = mean(hate_url$hate_crimes_combine),
                            sd = sd(hate_url$hate_crimes_combine)),
                col = "#1b98e0",
                size = 1) 
summary(hate_url$hate_crimes_combine)
```

Because we see the outliers, we would like to clear it. 

Now we can see the mean drop from 2.6460 to 2.4323, and max drop from 12.4758 to 5.4327 .


```{r echo=FALSE}
Q1 <- quantile(hate_url$hate_crimes_combine, .25)
Q3 <- quantile(hate_url$hate_crimes_combine, .75)
IQR <- IQR(hate_url$hate_crimes_combine)
new_hate_url <- subset(hate_url, hate_url$hate_crimes_combine> (Q1 - 1.5*IQR) & hate_url$hate_crimes_combine< (Q3 + 1.5*IQR))
ggplot(new_hate_url, aes(hate_crimes_combine)) +        
  geom_histogram(aes(y = ..density..),bins=20) +
  stat_function(fun = dnorm,
                args = list(mean = mean(new_hate_url$hate_crimes_combine),
                            sd = sd(new_hate_url$hate_crimes_combine)),
                col = "#1b98e0",
                size = 1) 
summary(new_hate_url$hate_crimes_combine)


```

After removing the outliner, let's compare the result of both group of higher than avg unemployed rate and lower than avg unemployed rate. There is more states has lower unemployed rate.

However, the states has lower unemployed rate tend to have a higher hate crime cases

```{r}
boxplot(new_hate_url$hate_crimes_combine ~ new_hate_url$high_unemployed, main = "Boxplot with hate crimes case of high unemployed rate", ylab = "hate crimes case", xlab = "high unemployed rate",col="blue")
```

```{r}
boxplot(new_hate_url$hate_crimes_combine ~ new_hate_url$high_housegoldincome, main = "Boxplot with hate crimes case of high household income", ylab = "hate crimes case", xlab = "high unemployed rate",col="blue")
```
```{r include=FALSE}

#education level and hate crime
summary(joined_tibble)

#find average high eduation level states
joined_tibble$high_education <-joined_tibble$PercentBachelorsOrHigher %g% 0.3128 

ggplot(joined_tibble, aes(x = high_education)) +
  geom_bar(fill="blue") +
  labs(
    x = "", y = "",
    title = "does the state have higher than avg high education rate?"
  ) +
  coord_flip() 

```
Also, the states has High education level tend to have a higher hate crime cases !!!???

```{r}

boxplot(joined_tibble$avg_hatecrimes_per_100k_fbi ~ joined_tibble$high_education, main = "Boxplot with hate crimes case of high education rate", ylab = "hate crimes case", xlab = "high education rate",col="blue")

```

### Inference

```{r}
hate_model <- lm(new_hate_url$hate_crimes_combine ~ new_hate_url$share_unemployed_seasonal_100k )
summary(hate_model)
hate_model2 <- lm(new_hate_url$hate_crimes_combine ~ new_hate_url$median_household_income )
summary(hate_model2)
```


the p-value is 0.6229 which mean it is not statistically significant. since the sample size is small and not huge, it should not affect the raito of N, so the high p-value is pretty accurate and meaningful.


also, one more thing bring my attention which is R-squared is 0.005542 which mean effect size is small, it also show in the plot which is far away from the regression line.

```{r echo=FALSE}
ggplot(data = new_hate_url, aes(x = share_unemployed_seasonal_100k, y = hate_crimes_combine)) +
  geom_point(colour = "blue", size = 1) +
  stat_smooth(method = "lm", se = FALSE, color= "red")
ggplot(data = hate_model, aes(x = .fitted, y = .resid)) +
  geom_point(colour = "blue", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color= "red") +
  xlab("Fitted values") +
  ylab("Residuals")
ggplot(data = new_hate_url, aes(x = median_household_income, y = hate_crimes_combine)) +
  geom_point(colour = "blue", size = 1) +
  stat_smooth(method = "lm", se = FALSE, color= "red")
ggplot(data = hate_model2, aes(x = .fitted, y = .resid)) +
  geom_point(colour = "blue", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color= "red") +
  xlab("Fitted values") +
  ylab("Residuals")
```


Also, the distribution is heavy tailed since there is few extreme prositve and negative residuals.


```{r}
ggplot(data = hate_model, aes(sample = .resid)) +
  stat_qq(colour = "blue", size = 1) + stat_qq_line(colour = "red", size = 1)
ggplot(data = hate_model2, aes(sample = .resid)) +
  stat_qq(colour = "blue", size = 1) + stat_qq_line(colour = "red", size = 1)
```
```{r  echo=FALSE}
m1 <- lm(PercentHighSchoolOrHigher ~ avg_hatecrimes_per_100k_fbi, data = joined_tibble)
summary(m1)

m2 <- lm(PercentBachelorsOrHigher ~ avg_hatecrimes_per_100k_fbi, data = joined_tibble)
summary(m2)

m3 <- lm(PercentHighSchoolOrHigher ~ hate_crimes_per_100k_splc, data = joined_tibble)
summary(m3)

m4 <- lm(PercentBachelorsOrHigher ~ hate_crimes_per_100k_splc, data = joined_tibble)
summary(m4)
```

```{r}
ggplot(data = joined_tibble, aes(x = avg_hatecrimes_per_100k_fbi, y = PercentHighSchoolOrHigher)) +
  geom_jitter() + 
  geom_smooth(method = "lm")

hist(m1$residuals)
qqnorm(m1$residuals)
qqline(m1$residuals)



ggplot(data = joined_tibble, aes(x = avg_hatecrimes_per_100k_fbi, y = PercentBachelorsOrHigher)) +
  geom_jitter() + 
  geom_smooth(method = "lm")

hist(m2$residuals)
qqnorm(m2$residuals)
qqline(m2$residuals)



ggplot(data = joined_tibble, aes(x = hate_crimes_per_100k_splc, y = PercentHighSchoolOrHigher)) +
  geom_jitter() + 
  geom_smooth(method = "lm")

hist(m3$residuals)
qqnorm(m3$residuals)
qqline(m3$residuals)



ggplot(data = joined_tibble, aes(x = hate_crimes_per_100k_splc, y = PercentBachelorsOrHigher)) +
  geom_jitter() + 
  geom_smooth(method = "lm")

hist(m4$residuals)
qqnorm(m4$residuals)
qqline(m4$residuals)
```





### Conclusion

Unlike our prediction that higher rates of hate crimes are tied to household income, unemployment rate and education level. Our analysis show that is not a strong relationship between unemployment rate, household income, education level and hate crimes case from all the plot and summary. There is too much variability in the model and High p value shows that the model is not statistically significant. 

We believe if the data break it down into zip code it may have a more accurate conclusion by comparing by zip code group within one state since they share more similar cultural values. This also help to lower other potential biases. It is more fair to pick a state for study instead of the whole US if we want to focus on the true relation hate crimes and unemployment.


https://fivethirtyeight.com/features/higher-rates-of-hate-crimes-are-tied-to-income-inequality/
https://www.census.gov/data/tables/2018/demo/education-attainment/cps-detailed-tables.html





